Reasonably Programmable Literal Notation

### 2018-06-22 icfp18/syntax-icfp18.tex
finished edits through 2.2.0

When TLM applications are nested, a distinct color can be used at each depth. For example, Fig.~\ref{fig:nesting-example} shows a program fragment where we transform an encoding of a chemical structure expressed using the standard SMILES notation for chemical structures \cite{anderson1987smiles}, extended with splicing notation, into a vector graphic, then embed this directly into a fragment of a webpage. We detail the TLM \li{$html at Html.t}, which implements HTML notation similar to that found in Ur/Web, in the accompanying technical report, and assume a TLM, \li{$smiles at Smiles.t} and a function, \li{Smiles.to_svg : Smiles.t => Html.t}, not shown. In Reason, \li{|>} is reverse function application.
\begin{figure}[t]
\begin{lstlisting}[numbers=none]
  $html `( SURL<div>
             <h3>Chemical Structure of Sucrose</h3>
             <$>EURL$smiles `(SCOLOR{ECOLORmono_glucoseSCOLOR}-O-{ECOLORmono_fructoseSCOLOR}ECOLOR)` |> Smiles.to_svgSURL</$>
           </div>EURL )`
\end{lstlisting}
\vspace{-2px}
\caption{A practical demonstration of nested TLM application. The colors communicate the segmentation.}
\label{fig:nesting-example}
\end{figure}

@@

Segment types can be communicated directly to the programmer upon request by an editor service For Reason, we are equipping the {Merlin} tool\todo{cite ICFP merlin paper}{} \cite{Merlin}, which is used by various Reason editor modes (e.g. for Emacs and Vim), with a new editor command that reports the expected type of the innermost spliced segment containing the cursor. Note that because the type is explicit, this information can be reported even when there is a parse or type error in a spliced expression. % An alternative design that simply reported the type assigned to each spliced expression based on the final expansion (which already exists in Merlin) would be less useful in situations where there is a type error in a spliced expression, or the programmer is in the process of editing the program and wants to know what type of expression to construct.

Segment types are somewhat analogous to the argument types of a function. The difference is that the argument signature of a function is the same every time the function is applied, i.e. it is associated with the function itself, whereas the inferred segmentation can differ for each choice of literal body. This, of course, gives TLMs substantially more notational flexibility, even relative to infix or mixfix function notation (where the number of subexpressions is fixed \cite{wieland2009parsing}).% (even infix or mixfix functions \cite{wieland2009parsing}).

@@

Let us turn now to the perspective of the TLM provider, whose principal task is to define the lexer and parser named in the TLM definition. In our implementation of TLMs for Reason, lexers are generated using \li{ocamllex} \cite{ocaml-manual} and parsers using Menhir \cite{pottier2016menhir}, which is a modern derivative of the parser generator Yacc~\cite{johnson1975yacc,TarditiDR:mly} with support for LR(1) grammars \cite{jourdan2012validating}. Reason itself is implemented using these same tools. 
We will say more about how lexer and parser definitions are resolved by our implementation in Sec. \ref{sec:implementation}. What is fundamentally important is that there is some clear protocol by which the lexer and parser names and the starting non-terminal name specified by the TLM definition, in this case \li{RegexLexer} and \li{RegexParser.start} on Lines 3-4 of Fig.~\ref{fig:regex-tlm-def}, are resolved to lexer and parser implementations that can be evaluated at expansion-time, i.e. during compilation.

The \li{ocamllex} and Menhir definitions used to implement \li{RegexLexer} and \li{RegexParser} are given in Fig.~\ref{fig:lexer-and-parser}. A fragment of the signature of the \li{Relit} library, which provides various necessary definitions and utilities to TLM providers, is shown in Fig.~\ref{fig:relit-util}. 
After describing this example in detail, we will outline some alternative parsing strategies.

@@

\subsubsection{Other Implementation Strategies}
Although \li{ocamllex} and Menhir are powerful tools, they are not right for every parsing job. For example, we might want to use a different parser generator, a parser combinator library \cite{Hutton1992d}, or post-process the result of calling an existing parser to produce a corresponding proto-expansion. Fortunately, it is easy to bypass \li{ocamllex} and Menhir: 
%The most obvious method is to define a trivial lexer that reads the entire literal body into a single token, and a trivial parser with one semantic action. In fact, this is not actually necessary because the system does not interface with \li{ocamllex} and Menhir directly. Rather, it is the job of the user's build system to make sure that by the time the compiler is invoked, these tools have already been invoked as necessary to produce ordinary OCaml modules satisfying the expected interface. It is just as well to implement this interface manually. 
the \li{Relit} library includes a trivial lexer, \li{Relit.TrivLexer}, and a few functors that take care of the necessary boilerplate in situations where we simply want to define the parser as a function of type \li{string => Relit.ProtoExpr.t}. For example, the accompanying technical report briefly describes how \li{$re_expr} can be expressed in terms of the Reason parser, a parse tree serializer and a tree transformation. It also details how the \li{$html} TLM shown previously is implemented using an existing, ``production grade'' HTML parsing library, \li{Markup.ml} \cite{markupml}, together with a simple post-processing step.

@@

\begin{figure}
\begin{subfigure}[t]{0.4\textwidth}
\begin{lstlisting}
  module Map = { 
    module type S = {
      type key;
      type t('a);
      let empty : t('a);
      let add : 
        key -> 'a -> t('a) -> t('a);
    /* ... */ }
  }
\end{lstlisting}
\vspace{-7px}
\caption{The \li{Map.S} signature}
\label{fig:map-sig}
\end{subfigure}
\hfill
\begin{subfigure}[t]{0.55\textwidth}
\begin{subfigure}[t]{\textwidth}
\begin{lstlisting}
  notation $map(M : Map.S, type a) at M.t(a) {
    lexer  MapLexer; parser MapParser; 
  }
\end{lstlisting}
\vspace{-6px}
\caption{The parametric TLM \li{$map}}
\vspace{5px}
\label{fig:map-tlm}
\end{subfigure}
\begin{subfigure}[b]{\textwidth}
\begin{lstlisting}
  notation $stringmap = $map(Misc.StringMap);
  let make_request id req v = $stringmap
    `( "session_id" SURL->EURL string_of_int(id)SURL,EURL
                req SURL->EURL v )`;
\end{lstlisting}
\vspace{-4px}
\caption{Applying \li{$map}}
\label{fig:map-ap}
\end{subfigure}
\end{subfigure}
\vspace{4px}
\caption{Literal notation for all finite map implementations using parametric TLMs.}
\vspace{-5px}
\label{fig:map-fig}
\end{figure}

@@

 The reader is directed to \emph{PFPL} \cite{pfple1} for a detailed introductory account of all of these language constructs. We will only tersely summarize the static and dynamic semantics of the core language because the particularities are not critical to our ideas. The static semantics is organized around the type formation judgement, $\istypeU{\Delta}{\tau}$, the expression typing judgement, $\hastypeU{\Delta}{\Gamma}{e}{\tau}$, and the pattern typing judgement, $\patType{\pctx}{p}{\tau}$. Type formation contexts, $\Delta$, track hypothesis of the form $\Dhyp{t}$, and typing contexts, $\Gamma$, track hypotheses of the form $x : \tau$. In the pattern typing judgement, $\pctx$ collects the typing hypotheses generated by $p$. These judgements are inductively defined in the accompanying technical reportal material along with necessary auxiliary structures and  standard lemmas. The dynamic semantics of $\miniVerseBase$ is organized around the judgements $\isvalU{e}$, which says that $e$ is a value, and $\evalU{e}{e'}$, which says that $e$ evaluates to the value $e'$. As in ML, evaluation can diverge (general recursion is possible via recursive types \cite{pfple1}). It can also result in match failure.%\emph{Type formation contexts}, $\Delta$, and \emph{typing contexts}, $\Gamma$ are defined in the accompanying technical report in the standard way \cite{pfpl}, 
% \[\begin{array}{ll}
% % \textbf{Judgement Form} & \textbf{Description}\\
% \istypeU{\Delta}{\tau} & \text{$\tau$ is a well-formed type}\\
% %\isctxU{\Delta}{\Gamma} & \text{$\Gamma$ is a well-formed typing context assuming $\Delta$}\\
% \hastypeU{\Delta}{\Gamma}{e}{\tau} & \text{$e$ is assigned type $\tau$}\\
% \ruleType{\Delta}{\Gamma}{r}{\tau}{\tau'} & \text{$r$ takes values of type $\tau$ to $\tau'$}\\
% \patType{\pctx}{p}{\tau} & \text{$p$ matches values of type $\tau$}\\
% & \text{and generates hypotheses $\pctx$} 
% \end{array}\]
% \emph{Type formation contexts}, $\Delta$, are finite sets of hypotheses of the form $\Dhyp{t}$. %Empty finite sets are written $\emptyset$, or omitted entirely within judgements, and non-empty finite sets are written as comma-separated finite sequences identified up to exchange and contraction. We write $\Delta, \Dhyp{t}$ when $\Dhyp{t} \notin \Delta$ for $\Delta$ extended with the hypothesis $\Dhyp{t}$. %Finite sets are written as finite sequences identified up to exchange.% We write $\Dcons{\Delta}{\Delta'}$ for the union of $\Delta$ and $\Delta'$.
% \emph{Typing contexts}, $\Gamma$, are finite functions that map each variable $x \in \domof{\Gamma}$, where $\domof{\Gamma}$ is a finite set of variables, to the hypothesis $\Ghyp{x}{\tau}$, for some $\tau$. The judgements above are inductively defined in the accompanying technical reportal material and validate standard lemmas, also given in the accompanying technical report.
%Empty typing contexts are written $\emptyset$, or omitted entirely within judgements, and non-empty typing contexts are written as finite sequences of hypotheses identified up to exchange and contraction. We write $\Gamma, \Ghyp{x}{\tau}$, when $x \notin \domof{\Gamma}$, for the extension of $\Gamma$ with a mapping from $x$ to $\Ghyp{x}{\tau}$, and $\Gcons{\Gamma}{\Gamma'}$ when $\domof{\Gamma} \cap \domof{\Gamma'} = \emptyset$ for the typing context mapping each $x \in \domof{\Gamma} \cup \domof{\Gamma'}$ to $x : \tau$ if $x : \tau \in \Gamma$ or $x : \tau \in \Gamma'$. % We write $\isctxU{\Delta}{\Gamma}$ if every type in $\Gamma$ is well-formed relative to $\Delta$.
% \begin{definition}[Typing Context Formation] \label{def:isctxU}
% $\isctxU{\Delta}{\Gamma}$ iff for each hypothesis $x : \tau \in \Gamma$, we have $\istypeU{\Delta}{\tau}$.
% \end{definition}

@@

The Relit preprocessor rewrites TLM applications by following the specification in the previous sections differing only in that the ``dependency variable'' is a ``dependency path'', here to \li{Notation_regex___relit.Dependencies}. The only difficult is in finding the TLM definition at all, which due to our integration into the module system may have been accessed indirectly, e.g. here via the \li{open} directive and in other situations via module synonyms, functors and so on. The trick is in the singleton restriction just described---if we can determine the signature for the module path that \li{Apply} is accessed from, we have the full definition in hand. To do so, we run the OCaml typechecker on this unexpanded representation. Because the TLM application raises an exception, its type can be generalized arbitrarily. As long as unrelated type errors at a particular depth are resolved (treating the TLM applications at that depth as ``holes'') expansion at that depth can proceed. This occasionally causes types that incidentally appear in error messages to be more general than the programmer might expect (e.g. here \li{restriction_template} will have type \li{'a -> 'b} during this ``pre-typing'' phase). 
%If this type ends up in an error message incidentally, e.g. due to a signature mismatch elsewhere, it may be somewhat confusing. 
We are working to improve error reporting and, more generally, improving integration with various other tools in the Reason / OCaml ecosystem.

>>>>

When TLM applications are nested, a distinct color can be used at each depth. For example, Fig.~\ref{fig:nesting-example} shows a program fragment where we transform an encoding of a chemical structure expressed using the standard SMILES notation for chemical structures \cite{anderson1987smiles}, extended with splicing notation, into a vector graphic, then embed this directly into a fragment of a webpage. We detail the TLM \li{$html at Html.t}, which implements HTML notation similar to that found in Ur/Web, in the accompanying technical report, and assume a TLM, \li{$smiles at Smiles.t} and a function, \li{Smiles.to_svg : Smiles.t => Html.t}, not shown. In Reason, \li{|>} is reverse function application.
\begin{figure}[t]
\vspace{-8px}
\begin{lstlisting}[numbers=none]
  $html `( SURL<div>
             <h3>Chemical Structure of Sucrose</h3>
             <$>EURL$smiles `(SCOLOR{ECOLORmono_glucoseSCOLOR}-O-{ECOLORmono_fructoseSCOLOR}ECOLOR)` |> Smiles.to_svgSURL</$>
           </div>EURL )`
\end{lstlisting}
\vspace{-2px}
\caption{A practical demonstration of nested TLM application. The colors communicate the segmentation.}
\label{fig:nesting-example}
\vspace{-4px}
\end{figure}

@@

Segment types can be communicated directly to the programmer upon request by an editor service For Reason, we plan to equip the {Merlin} tool \cite{Merlin}, which is used by various Reason editor extensions (e.g. for Emacs and Vim), with a new editor command that reports the expected type of the innermost spliced segment containing the cursor. Note that because the type is explicitly stated, this information can be reported even when there is a parse or type error in a spliced expression. % An alternative design that simply reported the type assigned to each spliced expression based on the final expansion (which already exists in Merlin) would be less useful in situations where there is a type error in a spliced expression, or the programmer is in the process of editing the program and wants to know what type of expression to construct.

Segment types are somewhat analogous to the argument types of a function. The difference is that the argument signature of a function is the same every time the function is applied, i.e. it is associated with the function itself, whereas the segmentation can differ for each choice of literal body. This, of course, is what gives TLMs substantially more notational flexibility, even relative to infix or mixfix function notation (where the number of subexpressions is fixed \cite{wieland2009parsing}).% (even infix or mixfix functions \cite{wieland2009parsing}).

@@

Let us turn now to the perspective of the TLM provider, whose principal task is to define the lexer and parser named in the TLM definition. Our implementation of TLMs for Reason, called \li{Relit}, assumes that the TLM provider is using \li{ocamllex} to generate the lexer \cite{ocaml-manual} and Menhir to generate the parser \cite{pottier2016menhir}. Menhir is a modern derivative of Yacc~\cite{johnson1975yacc,TarditiDR:mly} with support for LR(1) grammars \cite{jourdan2012validating}. Reason itself is implemented using these same tools. \li{Relit} requires that the TLM provider package the lexer and parser into an \li{ocamlfind} package named in the TLM definition---in the case of Fig.~\ref{fig:regex-tlm-def}, \li{regex_parser}---so that \li{Relit} can load them, together with their dependencies, at expansion-time. The TLM also specifies the name of the starting non-terminal---here, \li{start} on Line 4 of Fig.~\ref{fig:regex-tlm-def}. 
% What is fundamentally important is that there is some clear protocol by which the lexer and parser names and the starting non-terminal name specified by the TLM definition, in this case \li{RegexLexer} and \li{RegexParser.start} on Lines 3-4 of Fig.~\ref{fig:regex-tlm-def}, are resolved to lexer and parser implementations that can be evaluated at expansion-time, i.e. during compilation. 
The \li{ocamllex} and Menhir definitions used to implement \li{RegexLexer} and \li{RegexParser} appear in Fig.~\ref{fig:lexer-and-parser}. A fragment of the signature of the \li{Relit_helper} library, which provides various necessary definitions and utilities to TLM providers, appears in Fig.~\ref{fig:relit-util}. 
After describing this example in detail, we outline some alternative parsing strategies that do not directly require the use of \li{ocamllex} and Menhir in Sec.~\ref{sec:html-example}. In particular, we describe the \li{$html} TLM from Fig.~\ref{fig:nesting-example}.

@@

\subsubsection{Other Implementation Strategies}
\label{sec:html-example}
Although \li{ocamllex} and Menhir are powerful tools, they are not right for every parsing job. For example, we might want to use a different parser generator, a parser combinator library \cite{Hutton1992d}, or post-process the result of calling an existing parser to produce a corresponding proto-expansion. Fortunately, it is easy to bypass \li{ocamllex} and Menhir: 
%The most obvious method is to define a trivial lexer that reads the entire literal body into a single token, and a trivial parser with one semantic action. In fact, this is not actually necessary because the system does not interface with \li{ocamllex} and Menhir directly. Rather, it is the job of the user's build system to make sure that by the time the compiler is invoked, these tools have already been invoked as necessary to produce ordinary OCaml modules satisfying the expected interface. It is just as well to implement this interface manually. 
the \li{Relit} library includes a trivial lexer, \li{Relit.TrivLexer}, and a few functors that take care of the necessary boilerplate in situations where we simply want to define the parser as a function of type \li{string => Relit.ProtoExpr.t}. For example, the accompanying technical report briefly describes how \li{$re_expr} can be expressed in terms of the Reason parser, a parse tree serializer and a tree transformation. It also details how the \li{$html} TLM shown previously is implemented using an existing, ``production grade'' HTML parsing library, \li{Markup.ml} \cite{markupml}, together with a simple post-processing step.

@@

\begin{figure}
\begin{subfigure}[t]{0.4\textwidth}
\begin{lstlisting}
  module Map = { 
    module type S = {
      type key;
      type t('a);
      let empty : t('a);
      let add : 
        key => 'a => t('a) => t('a);
    /* ... */ }
  }
\end{lstlisting}
\vspace{-7px}
\caption{The \li{Map.S} signature}
\label{fig:map-sig}
\end{subfigure}
\hfill
\begin{subfigure}[t]{0.55\textwidth}
\begin{subfigure}[t]{\textwidth}
\begin{lstlisting}
  notation $map(M : Map.S, type a) at M.t(a) {
    lexer  MapLexer; parser MapParser; 
  }
\end{lstlisting}
\vspace{-6px}
\caption{The parametric TLM \li{$map}}
\vspace{5px}
\label{fig:map-tlm}
\end{subfigure}
\begin{subfigure}[b]{\textwidth}
\begin{lstlisting}
  notation $stringmap = $map(Misc.StringMap);
  let make_request id req v = $stringmap
    `( "session_id" SURL->EURL string_of_int(id)SURL,EURL
                req SURL->EURL v )`;
\end{lstlisting}
\vspace{-4px}
\caption{Applying \li{$map}}
\label{fig:map-ap}
\end{subfigure}
\end{subfigure}
\vspace{4px}
\caption{Literal notation for all finite map implementations using parametric TLMs.}
\vspace{-5px}
\label{fig:map-fig}
\end{figure}

@@

 The reader is directed to \emph{PFPL} \cite{pfple1} for a detailed introductory account of all of these language constructs. We will only tersely summarize the static and dynamic semantics of the core language because the particularities are not critical to our ideas. The static semantics is organized around the type formation judgement, $\istypeU{\Delta}{\tau}$, the expression typing judgement, $\hastypeU{\Delta}{\Gamma}{e}{\tau}$, and the pattern typing judgement, $\patType{\pctx}{p}{\tau}$. Type formation contexts, $\Delta$, track hypothesis of the form $\Dhyp{t}$, and typing contexts, $\Gamma$, track hypotheses of the form $x : \tau$. In the pattern typing judgement, $\pctx$ collects the typing hypotheses generated by $p$. These judgements are inductively defined in the accompanying technical report along with necessary auxiliary structures and standard lemmas. The dynamic semantics of $\miniVerseBase$ is organized around the judgements $\isvalU{e}$, which says that $e$ is a value, and $\evalU{e}{e'}$, which says that $e$ evaluates to the value $e'$. As in ML, evaluation can diverge (general recursion is possible via recursive types \cite{pfple1}). It can also result in match failure.%\emph{Type formation contexts}, $\Delta$, and \emph{typing contexts}, $\Gamma$ are defined in the accompanying technical report in the standard way \cite{pfpl}, 
% \[\begin{array}{ll}
% % \textbf{Judgement Form} & \textbf{Description}\\
% \istypeU{\Delta}{\tau} & \text{$\tau$ is a well-formed type}\\
% %\isctxU{\Delta}{\Gamma} & \text{$\Gamma$ is a well-formed typing context assuming $\Delta$}\\
% \hastypeU{\Delta}{\Gamma}{e}{\tau} & \text{$e$ is assigned type $\tau$}\\
% \ruleType{\Delta}{\Gamma}{r}{\tau}{\tau'} & \text{$r$ takes values of type $\tau$ to $\tau'$}\\
% \patType{\pctx}{p}{\tau} & \text{$p$ matches values of type $\tau$}\\
% & \text{and generates hypotheses $\pctx$} 
% \end{array}\]
% \emph{Type formation contexts}, $\Delta$, are finite sets of hypotheses of the form $\Dhyp{t}$. %Empty finite sets are written $\emptyset$, or omitted entirely within judgements, and non-empty finite sets are written as comma-separated finite sequences identified up to exchange and contraction. We write $\Delta, \Dhyp{t}$ when $\Dhyp{t} \notin \Delta$ for $\Delta$ extended with the hypothesis $\Dhyp{t}$. %Finite sets are written as finite sequences identified up to exchange.% We write $\Dcons{\Delta}{\Delta'}$ for the union of $\Delta$ and $\Delta'$.
% \emph{Typing contexts}, $\Gamma$, are finite functions that map each variable $x \in \domof{\Gamma}$, where $\domof{\Gamma}$ is a finite set of variables, to the hypothesis $\Ghyp{x}{\tau}$, for some $\tau$. The judgements above are inductively defined in the accompanying technical reportal material and validate standard lemmas, also given in the accompanying technical report.
%Empty typing contexts are written $\emptyset$, or omitted entirely within judgements, and non-empty typing contexts are written as finite sequences of hypotheses identified up to exchange and contraction. We write $\Gamma, \Ghyp{x}{\tau}$, when $x \notin \domof{\Gamma}$, for the extension of $\Gamma$ with a mapping from $x$ to $\Ghyp{x}{\tau}$, and $\Gcons{\Gamma}{\Gamma'}$ when $\domof{\Gamma} \cap \domof{\Gamma'} = \emptyset$ for the typing context mapping each $x \in \domof{\Gamma} \cup \domof{\Gamma'}$ to $x : \tau$ if $x : \tau \in \Gamma$ or $x : \tau \in \Gamma'$. % We write $\isctxU{\Delta}{\Gamma}$ if every type in $\Gamma$ is well-formed relative to $\Delta$.
% \begin{definition}[Typing Context Formation] \label{def:isctxU}
% $\isctxU{\Delta}{\Gamma}$ iff for each hypothesis $x : \tau \in \Gamma$, we have $\istypeU{\Delta}{\tau}$.
% \end{definition}

@@

The Relit preprocessor rewrites TLM applications by following the specification in the previous sections differing only in that the ``dependency variable'' is a ``dependency path'', here to \li{Notation_regex___relit.Dependencies}. The only difficult is in finding the TLM definition at all, which due to our integration into the module system may have been accessed indirectly, e.g. here via the \li{open} directive and in other situations via module synonyms, functors and so on. The trick is in the singleton restriction just described---if we can determine the signature for the module path that \li{Apply} is accessed from, we have the full definition in hand. To do so, we run the OCaml typechecker on this unexpanded representation. Because the TLM application raises an exception, its type can be generalized arbitrarily. As long as unrelated type errors at a particular depth are resolved (treating the TLM applications at that depth as ``holes'') expansion at that depth can proceed. This occasionally causes types that incidentally appear in error messages to be more general than the programmer might expect (e.g. here \li{restriction_template} will have type \li{'a => 'b} during this ``pre-typing'' phase). 
%If this type ends up in an error message incidentally, e.g. due to a signature mismatch elsewhere, it may be somewhat confusing. 
We are working to improve error reporting and, more generally, improving integration with various other tools in the Reason / OCaml ecosystem.
